---
title: "Assignment 9 Solutions: Other Outcomes"
subtitle: "Part 1: Ordinal and Multinomial Outcomes"
author: "Applied Quantitative Methods II, UC3M"
date: "Spring 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
geometry: margin=2cm
header-includes:
  - \renewcommand*\rmdefault{ppl}
  - \usepackage{setspace}\setstretch{1.2}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
```

```{r packages}
library(carData)
library(MASS)
library(nnet)
library(marginaleffects)
library(ggplot2)
data(BEPS)
```

# 1. Ordered logit: perceptions of the national economy

**a)** Explore the outcome and convert to an ordered factor:

```{r 1a}
table(BEPS$economic.cond.national)
BEPS$econ_ord = factor(BEPS$economic.cond.national, ordered = TRUE)
```

The distribution is concentrated in the middle categories (2, 3, and 4), with category 3 ("stayed about the same") being the modal response. Very few respondents chose the extreme ends (1 = got much worse, 5 = got much better). OLS treats the numeric values 1–5 as equally spaced, implying that the difference between "got much worse" and "got a little worse" is identical to the difference between "stayed the same" and "got a little better." This is almost certainly wrong for Likert-type survey items, where psychological distances between adjacent categories need not be equal. Ordered logit avoids this assumption by estimating threshold parameters that let the data determine the spacing of the latent scale.

**b)** Fit the ordered logit model:

```{r 1b}
m_ologit = polr(econ_ord ~ age + gender + Europe + political.knowledge,
                data = BEPS, Hess = TRUE)
summary(m_ologit)
```

Note that `polr()` uses a reversed sign convention: the raw coefficient reported in the output is negated relative to the standard parameterization. A positive raw coefficient in the `polr()` output corresponds to a *negative* association with higher ordinal categories. The raw coefficient on `Europe` is negative, which — after applying the sign reversal — implies a positive association: respondents with stronger pro-EU attitudes tend to perceive national economic conditions as having improved. This is plausible: Blair's government was broadly pro-European and pro-single-market, so EU supporters may have held more favorable views of its economic stewardship. For reliable interpretation of magnitudes, we use average marginal effects below.

**c)** Compute average marginal effects across all response categories:

```{r 1c}
avg_slopes(m_ologit)
```

The AMEs show the average change in the probability of each response category associated with a one-unit increase in each predictor. For `Europe`, the AMEs on the lower categories (1 and 2) are negative, while the AMEs on the higher categories (4 and 5) are positive, consistent with a positive association between pro-EU sentiment and more optimistic economic assessments. As a sanity check, the AMEs for any given predictor must sum to zero across the five categories because probabilities are constrained to sum to one.

**d)** Predicted probabilities by gender, at the mean of all other covariates:

```{r 1d}
predictions(m_ologit, newdata = datagrid(gender = c("female", "male")))
```

The predicted probabilities for the most pessimistic category (1 = got much worse) and the most optimistic category (5 = got much better) are shown separately for female and male respondents, with all other covariates held at their sample means. Any gender differences in these predicted probabilities should be modest, given that `gender` does not appear to be a strong driver of economic perceptions relative to the other covariates in the model. The overlapping confidence intervals for male and female respondents across most categories suggest that the gender gap in economic optimism is not large in this dataset.

# 2. Multinomial logit: vote choice

**a)** Set the reference category and fit the multinomial logit:

```{r 2a}
BEPS$vote = relevel(BEPS$vote, ref = "Conservative")
m_mlogit = multinom(vote ~ economic.cond.national + Blair + Hague +
                           Kennedy + Europe, data = BEPS, trace = FALSE)
summary(m_mlogit)
```

The model produces two sets of log-odds coefficients: Labour vs. Conservative and Liberal Democrat vs. Conservative. The coefficient on `Blair` in the Labour vs. Conservative equation is strongly positive: higher approval of Tony Blair is associated with substantially greater log-odds of voting Labour rather than Conservative. This makes intuitive sense — Blair was the Labour leader, so voters who rated him favorably were much more likely to have voted for his party. By contrast, the `Blair` coefficient in the Liberal Democrat vs. Conservative equation is expected to be smaller or near zero, since Blair approval does not strongly differentiate Liberal Democrat voters from Conservatives.

**b)** Compute average marginal effects across all predictors and outcome categories:

```{r 2b}
avg_slopes(m_mlogit)
```

The AME of `Blair` on the probability of voting Labour is positive and substantial. A one-unit increase in Blair approval (on the 1–5 scale) is associated with a meaningful increase in the average probability of voting Labour, holding all other variables constant. This reflects the strong personalization of vote choice in 1997: feelings toward the party leader were a major driver of vote intention, and Blair in particular was unusually popular relative to his Conservative counterpart.

**c)** The multinomial logit assumes Independence of Irrelevant Alternatives (IIA): the odds ratio between any two alternatives (e.g., Labour vs. Conservative) is unaffected by the presence or characteristics of the third alternative (Liberal Democrats). In the red bus / blue bus analogy, IIA fails because two alternatives are near-perfect substitutes and removing one simply shifts its probability to the other rather than distributing it proportionally. For British party choice, IIA is a moderate concern: Labour and the Liberal Democrats are both centre-left parties, sharing some ideological space, so some voters may treat them as partial substitutes in a way IIA cannot accommodate. The Conservatives, however, occupy a clearly distinct ideological position (right-wing), so the three-party menu is not as degenerate as two buses of different colours. Overall, IIA is plausible for Conservative vs. the others but is a more legitimate worry for the Labour/Liberal Democrat distinction.
