---
title: "Assignment 10 -- Solutions: Part 1 (Project Organization)"
subtitle: "Applied Quantitative Methods II, UC3M"
output:
  pdf_document:
    toc: false
geometry: margin=2cm
header-includes:
  - \renewcommand*\rmdefault{ppl}
  - \usepackage{setspace}\setstretch{1.2}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
```

This assignment is about project organization, not statistical analysis. The solutions below show the content of each file that students should create inside their `assignment10/` folder.

# 1. Folder structure

**a)** Create the folder structure from the terminal:

```{bash, eval=FALSE}
mkdir -p assignment10/data assignment10/analysis/output assignment10/plots/output
touch assignment10/Makefile assignment10/README.md
touch assignment10/analysis/models.R assignment10/plots/figures.R
```

**b)** Download the dataset and place it in `data/`. Students can download it manually from GitHub or use:

```{bash, eval=FALSE}
curl -L -o assignment10/data/corruption.dta \
  "https://raw.githubusercontent.com/franvillamil/AQM2/refs/heads/master/datasets/other/corruption.dta"
```

**c)** Example `README.md`:

```{markdown, eval=FALSE}
# Assignment 10: Corruption and Wealth Pipeline

This project builds a reproducible analysis pipeline examining the
relationship between GDP per capita and corruption across countries.

- `data/`: Contains the `corruption.dta` dataset (Transparency International
  Corruption Perceptions Index and UNDP GDP per capita).
- `analysis/`: R script that estimates regression models and saves a LaTeX table
  to `analysis/output/`.
- `plots/`: R script that produces a scatter plot saved to `plots/output/`.
- `Makefile`: Automates the full pipeline with `make`.
```

# 2. Analysis script

The file `analysis/models.R` should contain the following:

```{r, eval=FALSE}
# setwd("~/path/to/assignment10")
options(stringsAsFactors = FALSE)

library(haven)
library(modelsummary)

# ============================================================
# Load data
df = read_dta("data/corruption.dta")

# ============================================================
# Define constants
dep_var = "ti_cpi"
indep_var = "undp_gdp"

# ============================================================
# Clean data
df = df[!is.na(df[[dep_var]]) & !is.na(df[[indep_var]]), ]
cat("Observations:", nrow(df), "\n")

# Assertion
if(nrow(df) < 10) stop("Too few observations")

# ============================================================
# Estimate models
m1 = lm(ti_cpi ~ undp_gdp, data = df)
m2 = lm(ti_cpi ~ log(undp_gdp), data = df)

# ============================================================
# Save table
modelsummary(
  list("Level" = m1, "Log" = m2),
  output = "analysis/output/table_models.tex",
  stars = TRUE,
  gof_map = c("r.squared", "nobs"))

cat("Analysis complete. Table saved.\n")
```

We can verify that the script works by running it here (loading the data from the URL instead of a local path):

```{r}
library(haven)
library(modelsummary)

df = read_dta("https://raw.githubusercontent.com/franvillamil/AQM2/refs/heads/master/datasets/other/corruption.dta")

dep_var = "ti_cpi"
indep_var = "undp_gdp"

df = df[!is.na(df[[dep_var]]) & !is.na(df[[indep_var]]), ]
cat("Observations:", nrow(df), "\n")

if(nrow(df) < 10) stop("Too few observations")

m1 = lm(ti_cpi ~ undp_gdp, data = df)
m2 = lm(ti_cpi ~ log(undp_gdp), data = df)

modelsummary(
  list("Level" = m1, "Log" = m2),
  stars = TRUE,
  gof_map = c("r.squared", "nobs"),
  output = "markdown")
```

# 3. Plots script

The file `plots/figures.R` should contain:

```{r, eval=FALSE}
# setwd("~/path/to/assignment10")
options(stringsAsFactors = FALSE)

library(haven)
library(ggplot2)

# ============================================================
# Load and clean data
df = read_dta("data/corruption.dta")
df = df[!is.na(df$ti_cpi) & !is.na(df$undp_gdp), ]

# ============================================================
# Scatter plot: corruption vs log GDP
ggplot(df, aes(x = log(undp_gdp), y = ti_cpi)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "log(GDP per capita)",
    y = "Corruption Perceptions Index",
    title = "Corruption vs. GDP per capita") +
  theme_minimal()

ggsave("plots/output/scatter_corruption.pdf", width = 7, height = 5)

cat("Scatter plot saved.\n")
```

We can verify the plot here:

```{r, fig.height=4}
library(ggplot2)

ggplot(df, aes(x = log(undp_gdp), y = ti_cpi)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "log(GDP per capita)",
    y = "Corruption Perceptions Index",
    title = "Corruption vs. GDP per capita") +
  theme_minimal()
```

# 4. Makefile

The `Makefile` should contain:

```{makefile, eval=FALSE}
all: analysis/output/table_models.tex \
     plots/output/scatter_corruption.pdf

analysis/output/table_models.tex: analysis/models.R \
                                  data/corruption.dta
	Rscript --no-save analysis/models.R

plots/output/scatter_corruption.pdf: plots/figures.R \
                                     data/corruption.dta
	Rscript --no-save plots/figures.R
```

**a)** The `all:` rule lists both output files as targets. Running `make` without arguments builds everything.

**b)** The analysis rule specifies that `table_models.tex` depends on both the R script and the dataset. If either changes, `make` re-runs the recipe.

**c)** The plots rule follows the same pattern: the scatter plot depends on `figures.R` and the dataset.

**d)** To test, run `make` from the `assignment10/` directory in the terminal. Both scripts should execute, producing `analysis/output/table_models.tex` and `plots/output/scatter_corruption.pdf`. Running `make` a second time (without changes) should print "nothing to be done" because the outputs are already up to date.

**Important:** The recipe lines (the `Rscript` commands) must be indented with a **tab character**, not spaces. This is a Makefile requirement.
