\input{preamble.tex}

% Additional package for math
\usepackage{amsmath}

\begin{document}

\thispagestyle{empty}

\begin{center}
\textbf{\Large Problem Set 3: Applied Regression (II)}\\\vspace{10pt}
{\Large Applied Quantitative Methods for the Social Sciences II}\\\vspace{10pt}
Carlos III--Juan March Institute, Spring 2026
\end{center}

\vspace{10pt}
\noindent
\textbf{\large Instructions:}

\vspace{10pt}
\begin{itemize}
\setlength\itemsep{0pt}
  \item {\color{red}{\textbf{Deadline}}}: \textbf{February 26, before class}
  \item Submit your work as a \code{.R} file called \code{ps3.R} in your GitHub repository
  \item Use comments in your R code to answer conceptual questions and explain your analysis
  \item For visualizations, use the \code{marginaleffects} package
  \item You are encouraged to work together, but each person must submit their own code
\end{itemize}

\vspace{20pt}

\section{Conceptual Questions}

Answer these questions using comments in your R script.

\subsection{Question 1: Interaction Effects}

Consider the model: $Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 (X \times Z) + \varepsilon$

\begin{enumerate}[label=\alph*)]
    \item What is the marginal effect of $X$ on $Y$? Show how it depends on $Z$.
    \item Suppose $\beta_1 = 0.5$, $\beta_3 = -0.1$, and $Z$ ranges from 0 to 10. At what value of $Z$ does the effect of $X$ become zero?
    \item Why is it incorrect to interpret $\beta_1$ as ``the effect of $X$'' in this model?
    \item A researcher estimates this model and finds that $\beta_1$ is not statistically significant. They conclude that ``$X$ has no effect.'' Explain why this conclusion is problematic.
\end{enumerate}

\subsection{Question 2: Non-linear Relationships}

\begin{enumerate}[label=\alph*)]
    \item Consider the model $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \varepsilon$. What is the marginal effect of $X$ on $Y$? At what value of $X$ is the marginal effect zero?
    \item Suppose $\beta_1 = 2$ and $\beta_2 = -0.1$. Sketch the relationship between $X$ and $Y$. Is it U-shaped or inverted U-shaped?
    \item In the log-log model $\log(Y) = \beta_0 + \beta_1 \log(X) + \varepsilon$, what is the interpretation of $\beta_1$?
\end{enumerate}

\subsection{Question 3: Standard Errors and Inference}

\begin{enumerate}[label=\alph*)]
    \item Explain what heteroskedasticity is and why it affects standard errors but not point estimates.
    \item A researcher reports robust standard errors in their analysis. What does this mean? When should robust standard errors be used?
    \item Explain the difference between statistical significance and practical significance. Give an example where a result is statistically significant but not practically significant.
\end{enumerate}

\vspace{15pt}

\section{Applied Analysis: Gapminder Data}

For this problem set, you will use the Gapminder dataset, which contains country-level data on life expectancy, GDP per capita, and population over time.

\begin{lstlisting}
install.packages("gapminder")
library(gapminder)
data(gapminder)
\end{lstlisting}

You will also need the \code{marginaleffects} package for visualization:

\begin{lstlisting}
install.packages("marginaleffects")
library(marginaleffects)
\end{lstlisting}

\subsection{Question 4: Non-linear Relationships}

Focus on the year 2007 for this analysis.

\begin{enumerate}[label=\alph*)]
    \item Create a scatter plot of GDP per capita versus life expectancy. Based on the plot, comment on whether you think a linear model is appropriate.
    \item Estimate three models:
    \begin{itemize}
        \item Model 1: $\text{lifeExp} = \beta_0 + \beta_1 \cdot \text{gdpPercap} + \varepsilon$
        \item Model 2: $\text{lifeExp} = \beta_0 + \beta_1 \cdot \log(\text{gdpPercap}) + \varepsilon$
        \item Model 3: $\text{lifeExp} = \beta_0 + \beta_1 \cdot \text{gdpPercap} + \beta_2 \cdot \text{gdpPercap}^2 + \varepsilon$
    \end{itemize}
    Print summaries of the results.
    \item Compare the $R^2$ values of the three models. Which model fits the data best?
    \item Using Model 2, interpret the coefficient on $\log(\text{gdpPercap})$ in a comment. What happens to life expectancy when GDP per capita doubles?
\end{enumerate}

\subsection{Question 5: Interaction Effects}

Now we will examine whether the relationship between GDP and life expectancy varies by continent.

\begin{enumerate}[label=\alph*)]
    \item Estimate a model with $\log(\text{gdpPercap})$, continent, and their interaction:
    $$\text{lifeExp} = \beta_0 + \beta_1 \cdot \log(\text{gdpPercap}) + \beta_2 \cdot \text{continent} + \beta_3 \cdot \log(\text{gdpPercap}) \times \text{continent} + \varepsilon$$
    Print the results.
    \item Using the \code{marginaleffects} package, calculate the marginal effect of $\log(\text{gdpPercap})$ for each continent. Print the estimates and 95\% confidence intervals.
    \item Create a plot showing the predicted life expectancy across the range of GDP per capita for each continent. Include confidence bands. Save the plot.
    \item Interpret your findings substantively in a comment. Does the relationship between wealth and health differ by continent? What might explain these differences?
\end{enumerate}

\subsection{Question 6: Presenting Results}

\begin{enumerate}[label=\alph*)]
    \item Using your interaction model from Question 5, calculate the predicted life expectancy for:
    \begin{itemize}
        \item A poor African country (GDP per capita = \$1,000)
        \item A middle-income Asian country (GDP per capita = \$10,000)
        \item A wealthy European country (GDP per capita = \$40,000)
    \end{itemize}
    Print both point estimates and 95\% confidence intervals.
    \item Calculate the ``first difference'': How much higher is life expectancy in a wealthy European country compared to a poor African country? Print the 95\% confidence interval for this difference.
    \item Create a visualization that effectively communicates the key findings from your analysis. Save it and explain in a comment why you chose this particular visualization.
\end{enumerate}

\subsection{Question 7: Diagnostics}

\begin{enumerate}[label=\alph*)]
    \item Create a residuals vs. fitted values plot for your interaction model. Comment on whether there are any patterns that suggest model misspecification.
    \item Test for heteroskedasticity using the Breusch-Pagan test (use the \code{lmtest} package). Print the test statistic and p-value.
    \item Re-estimate your model using robust standard errors (use the \code{sandwich} or \code{estimatr} package). How do the confidence intervals change?
\end{enumerate}

\vspace{15pt}

\section{Synthesis Question}

\subsection{Question 8}

In approximately 300 words (as comments in your R script), discuss the importance of moving beyond simple regression tables when communicating research findings. Drawing on your analysis of the Gapminder data, explain:

\begin{itemize}
    \item Why predicted values and marginal effects are more informative than regression coefficients alone
    \item The value of visualizations in understanding complex relationships (like interactions)
    \item How to effectively communicate uncertainty in your estimates
\end{itemize}

\vspace{15pt}

\section{Submission}

Commit your \code{ps3.R} file to your GitHub repository before the deadline. Make sure your repository is public so I can access it.

Your R script should:
\begin{itemize}
  \item Be well-organized with clear section headers (using comments)
  \item Include all code needed to reproduce your analysis
  \item Include your answers to conceptual questions as comments
  \item Save any plots to files (e.g., using \code{ggsave()})
  \item Run without errors from top to bottom
\end{itemize}

\end{document}
